Supporting Material for:\\ Amplitude metrics for cellular \\ circadian bioluminescence reporters

### 2014-09-29 response_letter.tex

There has indeed been a great deal of literature concerning the accuracy of using the Gillespie stochastic simulation algorithm for non-elementary propensity functions.
In K. R. Sanft, {\itshape et al, IET Syst. Biol.} (2011), the authors conclude that the Michealis-Menten approximation overestimates the noise of the corresponding full system of equations.
In L. A. Widmer, {\itshape et al, J. Chem. Phys.} (2013), the authors prove that the noise of the full system is bounded on $[\nicefrac{3}{4}, 1]$ of the noise estimated by the Michealis-Menten approximation.
For Hill-type kinetics, the full system can take an arbitrary level of noise relative to the approximated system (L. A. Widmer, Master's Thesis, 2013).

The most important consideration is that converting the models to a full mass-action description of the system (or to a slightly less detailed total quasi-steady-state approximation as discussed in the literature cited by the reviewer) would require additional parameter values to be chosen.
This parameter set is not identifiable, as an infinite number of parameter choices exist that will yield equivalent continuous dynamics to the Hill approximation.
Each of these parameter sets will have different noise characteristics, which may not be reflected by the arbitrary noise characteristics arising from the Hill simplification.
However, since we do not specify these expanded parameters, we simply fit the noise characteristics of the approximated model (using the volume parameter) to yield physiologically realistic desynchronization rates.
Therefore, the validity of whether or not these approximated solutions closely represent the noise characteristics of the full solutions does not play a large role in determining the accuracy of the continuous approximation.

To address the larger point, we discuss reasons why our continuous approximations are less accurate for the larger model than the smaller model. 
We first point out that the method is an asymptotic approximation of a population of oscillators in the limit of no noise, as we account for the noise-induced desynchronization of oscillators but not noise-induced deviation from the limit cycle.
When noise is present, the `unperturbed' population consists of oscillators not only with a distribution of phases, but also with the states of each oscillator in a region of space near the deterministic limit cycle (this can be seen in the left plot of supplemental movies S1 \& S2).
As a result, the response to a pulse of an oscillator off the limit cycle can differ from the response of an on-limit oscillator with equivalent phase, decreasing the accuracy of the method for systems with high noise.
In practice, however, it takes a considerable amount of noise before the continuous predictions deviate substantially from the stochastic populations.
It is also important to consider that the larger model consists of many more state variables than the mass-action model, only one of which is shown in Fig.~S3.
Therefore, the difference in accuracy likely comes from increased deviation about the deterministic limit cycle in 8 spatial dimensions, rather than from non-elementary propensity functions.

### 2014-11-06 supplemental_info.tex
reflowing text in supplement
proofs recieved, some changes

\begin{document}

@@

\tablehead{\toprule Parameter & Value\\\midrule}

\begin{supertabular}{rl}
$vtp $  & 0.195   \\
$vtc1$  & 0.131   \\
$vtc2$  & 0.114   \\
$knp $  & 0.425   \\
$knc1$  & 0.259   \\
$vdp $  & 0.326   \\
$vdc1$  & 0.676   \\
\bottomrule\end{supertabular}\hfil
\begin{supertabular}{rl}
$vdc2$ & 0.608   \\
$kdp $ & 0.0115  \\
$kdc1$ & 1.15    \\
$vdP $ & 2.97    \\
$vdC1$ & 0.0338  \\
$vdC2$ & 1.52    \\
$vdCn$ & 1.69    \\
\bottomrule\end{supertabular}\hfil
\begin{supertabular}{rl}
$MC2n $ & 2.02    \\
$kdP  $ & 0.101   \\
$kdC1 $ & 3.32    \\
$kdCn $ & 0.0526  \\
$vaC1P$ & 0.0406  \\
$vdC1P$ & 0.00175 \\
$ktxnp$ & 3.0     \\
\bottomrule\end{supertabular}
\end{model}

>>>>

% Puts the affiliations on the cover page
\usepackage{authblk}

\begin{document}
\title{Supporting Material for:\\ Amplitude metrics for cellular \\ circadian bioluminescence reporters}
\author[1]{Peter C. St. John}
\author[2]{Stephanie R. Taylor}
\author[1]{John H. Abel}
\author[1,*]{Francis J. Doyle III}
\affil[1]{Department of Chemical Engineering, University of California Santa
Barbara, Santa Barbara, California 93106-5080}
\affil[2]{Department of Computer Science, Colby College, Waterville, Maine
04901}
\affil[*]{Email: \texttt{doyle@engineering.ucsb.edu}}
\date{}
\maketitle

@@

% \tablehead{}
\begin{tabular}{rl}
\toprule Parameter & Value\\\midrule
$vtp $  & 0.195   \\
$vtc1$  & 0.131   \\
$vtc2$  & 0.114   \\
$knp $  & 0.425   \\
$knc1$  & 0.259   \\
$vdp $  & 0.326   \\
$vdc1$  & 0.676   \\\bottomrule
\end{tabular}\hfil
%
\begin{tabular}{rl}
\toprule Parameter & Value\\\midrule
$vdc2$ & 0.608   \\
$kdp $ & 0.0115  \\
$kdc1$ & 1.15    \\
$vdP $ & 2.97    \\
$vdC1$ & 0.0338  \\
$vdC2$ & 1.52    \\
$vdCn$ & 1.69    \\\bottomrule
\end{tabular}\hfil
%
\begin{tabular}{rl}
\toprule Parameter & Value\\\midrule
$MC2n $ & 2.02    \\
$kdP  $ & 0.101   \\
$kdC1 $ & 3.32    \\
$kdCn $ & 0.0526  \\
$vaC1P$ & 0.0406  \\
$vdC1P$ & 0.00175 \\
$ktxnp$ & 3.0     \\\bottomrule
\end{tabular}
\end{model}
